Attrition and Achievement Gaps in Online Learning

### 2015-01-24 paper_v2.tex
updated version

Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (cf. Table \ref{tab:s1sum}). For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment. 
% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
On the initial course surveys, learners reported their gender, age, education, prior experience with the topic and MOOCs, enrollment intentions (on the OLEI scale \cite{kizilcec2015motivation}), and how many hours per week they intend to spend on the course. On the feedback survey, learners reported their satisfaction with the course (6-item index, range from 1 to 7, $M=5.29, SD=0.94, \alpha=.82$), and whether they `stopped participating in the course before it ended' or `remained active until the end'. Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (Figure \ref{fig:s1reason} lists reasons).\footnote{Instead of a select-all-that-apply question, learners were encouraged to consider each reason in turn by using binary-choice items. The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to mitigate acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region.

Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course. Behavioral indicators were chosen to represent course milestones: watching over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile. A regression analysis predicting course milestones and performance was performed based on a subset of learners who completed the initial course survey in the 17 MOOCs in our dataset. Out of the 63,620 learners who started the survey, we selected those with up to three missing values in the dataset, yielding a sample of $N=44,382$. Missing values were multiply imputed based on other survey responses and pooled estimates are reported, which account for the added uncertainty.\footnote{Multiple imputation and subsequent pooled analyses were performed with the $R$ mice package.}
% The data illustrated in Figure \ref{fig:s1reason} reflect the proportions of learners who selected each reason relative to all learners who reportedly stopped participating in each course.
% The correlation matrixes in Tables \ref{tab:s1reascor} and \ref{tab:s1cor2} are a linear combination of correlation matrixes for each course, assigning equal weight to all courses to discount for considerable variation in sample sizes.
% Information about age, gender, and education level were only available for learners who completed the initial course survey.
% We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior

\subsection{Results}

We address the research questions developed in the previous section in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across subsets of learners ($RQ2ab$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3ab$).

\subsubsection{Individual Differences in Attrition}

We fit a logistic mixed-effects (hierarchical) model to predict multiple binary persistence and performance metrics (milestones) to estimate individual differences in learner characteristics, including demographic and geographic features. Correlations between predictors were low ($|r|<0.16$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded. The coefficient on education, for instance, reflects the association between having a bachelor's degree and achieving a milestone, partialing out all other predictors, including age. Fourteen MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.\footnote{Data for 3 of the 17 MOOCs was unavailable due to anonymization.}

Figure \ref{fig:s1coefs} presents transformed regression coefficients with 95\% confidence intervals from each model. The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population. The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average aged male learners in North America without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it. In this sample of survey respondents, the baseline probability for each milestone was high: videos watched (27\% above 30\%, 16\% above 50\%, 8\% above 80\%) assessments attempted (78\% above 30\%, 55\% above 50\%, 33\% above 80\%), and grade (42\% above the 60$^\text{th}$, 31\% above the 80$^\text{th}$ percentile).

A number of significant and substantial individual differences emerged. Women were 10 to 15\% less likely than men to persist with lectures and assessments. Women, who constituted 41\% of learners in the sample, were also less likely to score a grade above the 60$^\text{th}$ percentile than men. Learners in Africa ($n = 2,097$), Asia ($n = 8,199$), and South America ($n = 2,574$) were 20 to 50\% less likely to persist in video watching and tended to achieve lower grades than learners in North America ($n = 18,353$). In contrast, learners in Europe ($n = 11,992$) exhibited 20 to 50\% higher persistence with lectures and assessemnts than learners in North America. While learners in Asia exhibited lower persistence with videos, they were more likely to attempt more assessments than learners from North America. Learners in Oceania ($n = 1,166$) performed similarly to those in North America and Europe.

Besides gender and geography, the older the learner, the more likely they were to persist in lectures and assessments, but the lower their relative performance. More educated learners, those with a bachelor's or higher degree, were somewhat more likely to start assessments, compared to those with less education, but were less likely to score highly on them. Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance. Learners who took prior MOOCs, however, were less likely to persist, yet achieved higher grades.

@@

\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Estimates from logistic mixed-effects model predicting persistence and performance milestones]{Estimates from logistic mixed-effects model predicting persistence and performance milestones. An estimate of -10 indicates a 10\% lower chance of achieving a milestone.\label{fig:s1coefs}}
\end{figure*}

@@

\subsubsection{Learner Satisfaction}
% RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
% RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?

Attrition can be a sign of success if a learner has achieved their personal goal. We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who participanted to different degrees ($RQ2b$). Satisfaction ratings were compared with a mixed-effects regression to model differences across courses and generalize to a larger population of MOOCs. Learners who reported that they stopped participating were 9\% less satisfied than those who reported not to have stopped ($t_{15140}=18, p<.001$). Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied. While the the proportion of videos significantly predicted higher satisfaction ($t_{15700}=14, p<.001$), the effect was relatively weak. Those who watched over 75\% of videos were only 5\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=4.97, M_{>\text{75\%}}=5.22$).

% We also examined differences in satisfaction between learners grouped by certain behaviors. First, we classified learners into 4 groups based on the fraction of course videos they viewed. The box plots for the satisfaction scores in each group are shown in figure \ref{fig:study1_satsfaction_by_vid_groups}. The mean satisfaction score increases slightly as the number of videos viewed by the learners increases, with the exception of the scores reported by learners in the 0-25\% group. Possible reasons for this are learners who join only to view a small part of the course or learners who join merely to explore.

@@

% Say how many actually reported N=?
Learners' reasons for disengaging from a MOOC were examined using survey responses ($RQ3a$). Across 17 MOOCs, learners selected reasons for disengaging from a pre-defined list. This list may not be collectively exhaustive, which is addressed in Study 2. The distribution of chosen reasons is illustrated in Figure \ref{fig:s1reason}. Around half of the respondents indicated two time-related reasons as influencing their decision to stop participating. Learners faced issues keeping up with deadlines and found that the course demanded too much time. Other reasons for attrition were less prevalent, but still reported by around 10 to 25\% of respondents. Notably, just under 20\% of respondents in a typical course (median proportion) stopped participating because they learned all they intended to learn.

% say how this was done (linear combination of correlation matrices; exluding any? if so, N=?)
To investigate the underlying factor structure of the reported reasons for disengaging, we conducted an exploratory factor analysis. The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot. The following five factors emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines); second, {\em late start} (starting course late); third, {\em difficulty} (too advanced, exam too difficult); fourth, {\em format \& content} (didn't enjoy online format, confusing videos, materials not explained well), and fifth, {\em goals \& expectations} (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable). The technical difficulties item failed to load onto any factor.

% NEED TO SAY THAT IT'S LITTLE DATA HERE (n=?)
We expected reasons for disengaging from the course to vary by learner characteristics ($RQ3b$), such as learners' motivations for enrolling in the course. A correlation matrix of learners' enrollment intentions and their subsequent reasons for disengaging revealed a small number of notable associations. Taking the course with friends or colleagues was correlated with disengaging because the course was not challenging enough, the material was too advanced for the learner's background, and the credential was not valuable enough to the learner ($r=.24, .20, .16$, respectively). Not surprisingly, learners who enrolled to experience an online course tended to stop participating because they were only exploring the course ($r=.17$). Learners looking to improve their English tended to experience technical difficulties ($r=.17$), and those who enrolled for personal growth and enrichment tended to struggle to keep up with deadlines ($r=.16$).

% other covariates: e.g. do we see geolocation difference for language and tech?
% also gender differences?

@@

\includegraphics[width=\maxwidth]{figure/s1reason} \caption[Proportion of learners who indicated each reason for attrition across 17 MOOCs]{Proportion of learners who indicated each reason for attrition across 17 MOOCs\label{fig:s1reason}}
\end{figure*}

@@

\subsection{Discussion}
% need to adjust once the rest is finalized above
A number of striking findings emerged from the preceeding analyses. We found strong evidence for gender inequalities in attrition and, to some extent, performance. Women were less likely to persist, even when adding a number of other demographic, geographic, and preference indicators to the model, which highlights the robustness of the relationship. The MOOCs in our dataset were on diverse topics, including some on topics in the humanities and business with high rates of participation from women. Besides gender differences, attrition was substantially higher for learners from Africa, Asia, and South America, compared to North America, Oceania, and Europe, in terms of lectures watched. We also identified a large achievement gap in terms of grades between learners from these continents. The current generation of MOOCs is seemingly not immune to inequalities that have pervaded traditional educational settings. It remains unclear, however, to what extent these gaps result from differences in Internet access and language barriers, or potentially, from experiences of psychological threat to learners' identity or sense of social belonging \cite{walton2007question}.

Levels of self-reported satisfaction were generally high and did not vary much by level of persistence, although learners who reportedly disengaged from the course tended to be less satisfied than those who reportedly persisted until the end. Learners' reasons for disengaging from the course clustered into five key factors: general time issues; problems due to starting the course late; course difficulty; issues with the course format or content, and a mismatch or fulfillment of expectations. General time issues were overall the most pronounced. Attrition in community colleges has been interpreted as a sign of success in prior work. This view resonates in the MOOC setting, where around 20\% of learners who disengaged reportedly had achieved their goal, and where progress in the course and learner satisfaction were only weakly related.

Learners may disengage from the course when they become too occupied by higher-priority tasks. Enrolling in a MOOC neither guarantees that a learner can or intends to allocate enough time to go through the course. According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gain. A low capacity for volitional control \cite{corno2001volitional} could lead learners to misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task. This implies that learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control.

@@

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. This study investigates reasons for attrition based on psychological measures and learners' own descriptions of the challenges they faced. We invited learners who were likely to drop out of a course to provide feedback. These learners were identified with a predictive model and invited to complete a survey. The survey included self-report measures of psychological constructs to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners. Moreover, most learners indicated in Study 1 that they disengaged due to a shortage of time. A variety of factors could influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal committments are presumably large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}. Low volitional control could reduce learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (e.g., watching TV). We constructed the following hypotheses about the role of volition in attrition due to a shortage of time:

@@

Two research assistants independently developed codebooks for the resulting 448 non-empty open responses. Their codebooks were consolidated and applied on Amazon Mechanical Turk (AMT) in three iterations. The codebook was updated in each iteration to reliably fit the open response data. Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement. In the first iteration, 250 randomly selected responses were each coded by four `classification experts,' which is an AMT-specific qualification. In the second iteration, the remaining responses were coded in the same way. In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see !!!!!!![URL].} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.

% Learners also rated the extent to which their progress was hindered by a number of obstacles. 
Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adpated from \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).

\subsection{Results}

@@

The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course. While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e. exhibit lower volitional control). These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$). Reasons for learner attrition were measured using iteratively coded open responses. Table \ref{tab:s2reas} provides each coded reason and percentage of open responses that indicated it. A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$). Hence, if a learner had not enough time, this tended to be the primary challenge.

@@

@@

This study of over 60,000 online learners, combining behavioral and survey data, extends the empirical base of research on attrition in online learning environments. The three major findings are, \emph{first}, the presence of substantial gaps in persistence and performance between learners with different demographic features and from different geographic locations; \emph{second}, the relation between personal success in online learning (by learners' own judgment) and higher levels of goal-striving, social belonging, and growth mindset, and \emph{third}, the challenges faced by online learners, with the shortage of time as the major obstacle, which was frequently related to learners' low capacity for volitional control. Moreover, we developed a survey item for measuring reasons for disengaging from a course (cf. Table \ref{tab:s2reas}), which should be tested and potentially extended based on more empirical evidence from additional courses.

@@

\section{Acknowledgments}
We thank Candace Thille, Emily Schneider, and Andreas Paepcke for their helpful feedback, and Elise Ogle and Ruth Bram for helping with the development of the codebook. The authors contributed equally to this work.

>>>>

Twenty MOOCs on topics in a variety of disciplines were selected for investigation. Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and a subset of behavioral indicators for each course. For each course, we combined data from an initial course survey (distributed in the first weeks of the course), a feedback survey (distributed in the last weeks of the course), and behavioral data collected in the learning environment. On the initial course survey, learners reported their gender, age, education, prior experience with the topic and MOOCs, and how many hours per week they intend to spend on the course. On the feedback survey, learners reported their satisfaction with the course (5-item index, possible range from 0 to 5.2, $M=3.95, SD=0.81, \alpha=.81$), and whether they `stopped participating in the course before it ended' or `remained active until the end'. Those who indicated that they stopped participating also reported `what influenced their decision to stop participating' using a grid of 14 binary-choice items (cf. Figure \ref{fig:s1reason}).\footnote{Instead of a select-all-that-apply question, this question style encouraged learners to consider each reason separately. The response options were ``Did influence my decision to stop" and ``Did NOT influence my decision to stop" to reduce acquiescence bias relative to a Yes/No or Agree/Disagree scale.} Geographic location was determined by IP address and aggregated by continental region. Behavioral indicators were chosen to represent course milestones: watching (or downloading) over 30\%, 50\%, 80\% of lecture videos, attempting over 30\%, 50\%, 80\% of assessments, and achieving a grade above the 60$^\text{th}$, 80$^\text{th}$ percentile.

To analyze individual differences, we fit a logistic mixed-effects (hierarchical) model predicting binary persistence and performance milestones using learner characteristics, including demographic and geographic features. The analysis included 67,333 learners who completed the initial course survey, but excluded 381 learners who left all questions unanswered and learners in courses C6, C7, C19, C20 due to missing survey items. Any remaining missing survey responses were multiply imputed using other survey responses. We report pooled estimates that account for added uncertainty.\footnote{Multiple imputation ($m=10$) using the $R$ mice package.} Correlations between predictors were low ($|r|<0.15$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded. Courses were modeled as a random effect (instead of including dummy variables) to generalize to a larger population of similar MOOCs.

\subsection{Results and Discussion}

We address the posed research questions in three parts: first, a regression analysis of differential attrition and performance ($RQ1$); second, a comparison of satisfaction levels across learner subgroups ($RQ2$), and third, a descriptive analysis of reported reasons for disengaging from the course ($RQ3$).

\subsubsection{Individual Differences in Attrition and Performance}

Individual differences in attrition and performance are illustrated in Figure \ref{fig:s1coefs} as transformed regression coefficients with 95\% confidence intervals for each milestone. The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population. The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average-aged male learners in Northern America\footnote{Learners in the United States, Canada, Bermuda, Greenland.} without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it. The baseline probability for each milestone was relatively high: lectures watched (20\% above 30\%, 11\% above 50\%, 4\% above 80\%), assessments attempted (37\% above 30\%, 25\% above 50\%, 18\% above 80\%), and grades (30\% above the 60$^\text{th}$, 22\% above the 80$^\text{th}$ percentile).

A number of significant and substantial individual differences emerged. Women were 12 to 20\% less likely than men to persist with lectures and assessments. Women, who constituted 34\% of learners in the sample, were also 10\% (7\%) less likely than men to score a grade above the 60$^\text{th}$ (80$^\text{th}$) percentile.  Moreover, learners in Africa ($N = 2,458$), Asia ($N = 13,280$), and Latin America ($N = 5,344$) were 24 to 50\% less likely to persist with lectures and 31 to 52\% less likely to achieve grade milestones than learners in Northern America ($N = 26,651$). In contrast, learners in Europe ($N = 17,988$) exhibited 16 to 30\% higher persistence with lectures and assessemnts than learners in Northern America. While learners in Asia exhibited lower persistence with videos, they were at least as likely to attempt assessments as learners in North America. Learners in Oceania ($N = 1,612$) performed similarly to those in Northern America and Europe. Besides gender and geography, older learner were more likely to persist with lectures and assessments, but achieved lower grades than younger learners. More educated learners, those with a bachelor's or higher degree, were 15 to 25\% more likely to persist and 10\% more likely to score above the 60$^\text{th}$ percentile. Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance. Learners who participated in prior MOOCs had mixed persistence patterns, but achieved higher grades.

The large scale of MOOCs has revealed a global achievement gap and a gender achievement gap in online learning. Performance and persistence in the course was substantially lower for women and learners in Africa, Asia, and Latin America than men and learners in Europe, Oceania, and Northern America. There was a good representation from each geographic region in the dataset and the courses under consideration were on diverse topics, including some in the humanities and business with high female participation rates. The strength and significance of these relationships is remarkable, in particular with several additional demographic, geographic, and preference measures in the model. The first generation of large-scale online learning, characterized by MOOCs, is apparently not immune to inequalities that have pervaded traditional educational settings. The present work cannot move beyond presenting evidence for the existence of these gaps and discussing potential causes. Of course there exists no innate feature of women or people in certain locations that is responsible for these gaps; but rather, gender and location are associated with countless individual characteristics, some of which impact academic achievement. The achievement gaps could plausibly result from differences in Internet access, language barriers, or from feelings of psychological threat, such as fears of confirming a negative stereotype \cite{spencer1999stereotype} or not belonging in the course \cite{walton2007question}. More work is needed to identify potential causal mechanisms underlying these gaps.

@@

\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Individual differnces in persistence and course grades showing substantial achievement gaps]{Individual differnces in persistence and course grades showing substantial achievement gaps. An estimate of -10 indicates a 10\% lower chance of achieving a milestone.\label{fig:s1coefs}}
\end{figure*}

@@

\subsubsection{Satisfaction}
% RQ2a: What proportion of learners who disengaged were satisfied with their achievements in the course?
% RQ2b: How does learners' behavior in the MOOC vary by their level of satisfaction?

Attrition can be a sign of success if a learner has achieved their personal learning objectives. We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who persisted to different extents ($RQ2b$). Satisfaction ratings were compared using mixed-effects regression ($N=23,390$). Learners who reported that they stopped participating were 12\% less satisfied than those who reported not to have stopped ($t_{453}=21, p<.001$). Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied. The proportion of videos significantly predicted higher satisfaction ($t_{101}=20, p<.001$), such that those who watched over 75\% of videos were 8\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=3.80, M_{>\text{75\%}}=4.09$). As levels of self-reported satisfaction were generally high, it remains unclear if the estimated differences in satisfaction are consequential. Learners who disengaged early were only somewhat less satisfied than those who persisted in the course, which suggests that some learners arrived with modest course goals and achieved them. The following analysis of reasons for attrition can shed more light on this hypothesis.

@@

A fraction of learners who completed the course feedback survey reported having left the course early and indicated reasons for disengaging on a pre-defined list ($RQ3a$). This list may not be collectively exhaustive, which is addressed in Study 2, but responses were collected from 1,698 learners in 20 different courses. Figure \ref{fig:s1reason} illustrates how frequently each reason was selected across courses. Around half of the respondents indicated two time-related reasons as influencing their decision to stop participating. In a typical course (median proportion), 66\% faced issues keeping up with deadlines and for 46\% the course demanded too much time. According to expectancy-value theory, the learner is a rational agent who allocates time to different activities to maximize subjective gains. The shortage of time could be due to an abundance of higher-priority tasks, as enrolling in a MOOC neither guarantees that a learner can nor intends to allocate enough time to complete the course. Alternatively, a low capacity for volitional control \cite{corno2001volitional} could lead learners to misallocate time to lower-priority activities that promise nearer gratification, instead of allocating it to the next highest priority task. Learners' general time issues may be a product of high-priority obligations as well as a result of a low capacity for volitional control. We investigate this hypothesis in Study 2.

Other reasons for attrition were less prevalent, but still reported by around 10 to 25\% of respondents. Notably, 17\% of respondents in a typical course stopped participating because they learned all they intended to learn. This finding resonates with some prior work on attrition in community colleges, where attrition has been interpreted as a sign of success. It also supports the observation that progress in the course and satisfaction were only weakly related. We investigated the underlying factor structure of the reported reasons for disengaging using an exploratory factor analysis. The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot. The following four-factor solution emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines);  second, {\em difficulty} (too advanced, exam too difficult); third, {\em format \& content} (didn't enjoy online format, confusing videos, materials not explained well, not what I'm looking for, technical difficulties), and fourth, {\em goals \& expectations} (only exploring, learned all I wanted, not challenging, credential not valuable). The `late start' item failed to load onto any factor. General time issues and difficulty are very interpretable factors, while the other two factors comprise a complex set of reasons that should probably be further separated.

Reasons for disengaging from a course are likely to vary by learner characteristics ($RQ3b$). To explore if the substantial individual differences in attrition identified above could be associated with different reasons for disengaging, we investigated variation by geographic region ($N = 1,698$) and gender ($N = 1,039$ due to missing gender information). Learners in Africa, Asia, and Latin America were more likely to disengage due to technical difficulties (61\% more likely), too demanding deadlines, and starting late ($p\text{s} < .03$), while learners in Europe, Oceania, and Northern America were more likely to disengage because the course was not a good fit for them ($p=.02$). Women were more likely than men to report disengaging due to technical difficulties and because the course required too much time ($p<.01$). Technical difficulties were a critical individual difference that could be a source of the observed achievement gaps. The pre-defined list of reasons for attrition unfortunately did not include an item on language difficulties, which could be another plausible source of the global achievement gap.

@@

\includegraphics[width=\maxwidth]{figure/s1reason} \caption[Distribution of self-reported reasons for attrition across 20 MOOCs]{Distribution of self-reported reasons for attrition across 20 MOOCs.\label{fig:s1reason}}
\end{figure*}

@@

@@

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. This study investigates reasons for attrition based on psychological measures and learners' own descriptions of the challenges they faced. We collected feedback from learners who were likely to have stopped following the course. They were identified with a predictive model and invited to complete a survey. The survey included self-report measures of psychological constructs to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of online learners. Moreover, in Study 1, most learners indicated that they disengaged due to a shortage of time. A variety of factors could influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal committments are presumably large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is a learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}. Low volitional control could reduce learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (watching TV, etc.). We constructed the following hypotheses about the role of volition in time-related attrition:

@@

Two research assistants independently developed codebooks for the resulting 448 non-empty open responses. Their codebooks were consolidated and applied on Amazon Mechanical Turk (AMT) in three iterations. The codebook was updated in each iteration to reliably fit the open response data. Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement. In the first iteration, 250 randomly selected responses were each coded by four `classification experts,' which is an AMT-specific qualification. In the second iteration, the remaining responses were coded in the same way. In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see http://kizilcec.com/rsc/las2015attrition.pdf.} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.

Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adapted from http://mindsetonline.com/testyourmindset}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).

\subsection{Results and Discussion}

@@

The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course. While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e. exhibit lower volitional control). These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$). Reasons for learner attrition were measured using iteratively coded open responses. Table \ref{tab:s2reas} provides each coded reason and percentage of open responses that indicated it. This list of reasons maps partially onto the factors identified in Study 1 (general time issues, difficulty, format \& content, and goals \& expectations). A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$). Hence, if a learner had not enough time, this tended to be the primary challenge.

@@

The majority of learners indicated that they had not enough time: some struggled due to higher priority obligations that took up too much time, while others were hindered by low levels of volitional control ($H4a$). A learner needs to be highly self-regulated to be successful in an open learning environment, such as a MOOC. Very few learners explicitly indicated issues with volitional control in their open responses. They may not possess the metacognitive skills to consciously notice this issue or they preferred not to admit to this undesirable trait. Learners who reported a shortage of time without providing specific reasons were more likely to have insufficient volitional control ($H4b$). This could be a useful proxy for identifying issues with volitional control in future work. Overall, the results suggest that learners' capacity for volitional control can be a critical ingredient for learners' success and warrants further investigation \cite{corno2001volitional}.

@@

This study of over 100,000 online learners in 21 courses, combining behavioral and survey data, extends the empirical base of research on attrition in online learning environments. The three major findings are, \emph{first}, the presence of substantial gaps in persistence and performance between learners of different gender and from different geographic locations; \emph{second}, the relation between personal success in online learning (by learners' own judgment) and higher levels of goal-striving, social belonging, and growth mindset, and \emph{third}, the challenges faced by online learners, with the shortage of time as the major obstacle, which was frequently related to learners' low capacity for volitional control. Moreover, we developed a survey item for measuring reasons for disengaging from a course (cf. Table \ref{tab:s2reas}), which should be tested and potentially extended based on more empirical evidence from additional courses.

@@

\section{Acknowledgments}
We thank Candace Thille, Emily Schneider, and Andreas Paepcke for their helpful feedback, and Elise Ogle and Ruth Bram for their assistance developing a codebook. The authors contributed equally to this work.
